k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-14 paper.Rmd
outline

Formally, anticlustering is used to partition $N$ objects into $K$ exhaustive and mutually exclusive subgroups $C = (C_1, \ldots, C_K)$. The partitioning restrictions are formalized as follows:

@@

@brusco2019 outlined the importance of simultaneously considering multiple criteria when addressing anticlustering problems. In particular, they argued that anticlustering algorithms should incorporate an objective of between-group similarity---such as the diversity---and pairwise within-group dissimilarity---i.e., the dispersion---at the same time. To this end, @brusco2019 presented a bicriterion algorithm that simultaneously maximizes the diversity $f_3$ and the dispersion $f_5$. While in the next section, we will discuss their approach in more detail. We endorse Brusco et al.'s bicriterion approach for anticlustering, and advance it in several ways in the current paper. Specifically, we offer several contributions: First, we allow that not only the standard diversity $f_3$ can be optimized alongside the dispersion, but instead we allow that the k-means, k-plus or the average diversity can be used as measures of between-group similarity (i.e., the criteria $f_1$, $f_2$ and $f_4$). Second, we allow that the dispersion and $f_1$ -- $f_4$ can be optimized using different dissimilarity matrices. This is useful if both criteria should be based on different information, or to induce custom cannot-link constraints, as we will show. As out most important contribution, we combine a novel exact algorithm for maximum dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. Despite its theoretical computational hardness, our optimal method scales to rather large data sets. It generates multiple partitions that we use as initializations of the BILS's local search, thus guaranteeing that the BILS also returns a partition with optimal dispersion. In a simulation study, we compared several adaptations of the BILS.

>>>>

Formally, anticlustering is used to partition $N$ objects into $K$ exhaustive and mutually exclusive subsets $C = (C_1, \ldots, C_K)$. The partitioning restrictions are formalized as follows:

@@

@brusco2019 outlined the importance of simultaneously considering multiple criteria when addressing anticlustering problems. In particular, they argued that anticlustering algorithms should incorporate an objective of between-group similarity---such as the diversity---and pairwise within-group dissimilarity---i.e., the dispersion---at the same time. To this end, @brusco2019 presented a bicriterion algorithm that simultaneously maximizes the diversity $f_3$ and the dispersion $f_5$. We endorse Brusco et al.'s bicriterion approach for anticlustering, and advance it in several ways in the current paper. First, we allow that not only the diversity $f_3$ can be optimized alongside the dispersion, but instead we allow that the k-means, k-plus or the average diversity can be used as measures of between-group similarity (i.e., the criteria $f_1$, $f_2$ and $f_4$). Second, we allow that the dispersion can be optimized on the basis of a different dissimilarity matrix than the diversity (or a different objective of between-group similarity). This is useful if the criteria should be based on different information, or to induce custom cannot-link constraints, as we will show. As our most important contribution, we combine a novel exact algorithm for maximum dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. Our paper is organized as follows. In the next section, we review the bicriterion method by @brusco2019 in detail. After that, we provide the technical background for our extensions. We then provide several evaluations of our contributions: (a) we probe the practical feasibility of our optimal algorithm for maximum dispersion, (b) we compare different adaptations of the BILS with regard to their ability to providing high diversity on top of an optimal dispersion, and (c) we provide practical examples on using our new methods.
